---
title: "Modeling plant stable isotopes using the 3PG forest growth model in R"
author: "Daniel M Griffith"
date: "April 12, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The r3PG R package implements the 3-PG (Physiological Processes Predicting Growth) model which was originally developed by [Landsberg and Waring (1997)](https://www.sciencedirect.com/science/article/pii/S0378112797000261). Specifically, this R implementation of the model is based directly on the Python version created by [Wei et al. (2014)](https://www.sciencedirect.com/science/article/pii/S0378112713007226) to incorporate predictions of stable isotope compositions. This vignette is focused specifically on using r3PG and predicting carbon and oxygen isotopic values. For a thorough introduction to 3PG and raw data for many of the examples presented here, check out the [course materials](http://3pg.forestry.ubc.ca/3-pg-course/) available online.

As implemented in R, the 3PG model requires two types of input which are climate data and information about the site and tree species. This structure is a legacy from the earlier vesions, and for compatibility ".cfg" files used in the Python version can also be used. However, r3PG also accepts native R objects and includes functions for reading and editing model configuration data within R. 

Below, we provide a workflow for comparing the stable isotope predictions for a variety of tree species at four sites with different climates.

## Installation

Install r3PG from GitHub using devtools.

```{r installation, message = FALSE, warning = FALSE, results = 'hide'}
install.packages("devtools", repos = "http://cran.us.r-project.org")
library(devtools)
install_github(repo = "griffithdan/r3PG")
library(r3PG)
```

## Load example climate data

The "r3PG" R package has monthly climate normals available for four example sites.

```{r, results='hide', fig.align = 'center', fig.width = 8, fig.height = 5}
data("clim.WesternOR")
data("clim.Argentina")
data("clim.NorthCarolina")
data("clim.BritishColumbia")

par(mfrow = c(1,2))
  with(clim.WesternOR, plot(Tav ~ c(1:12), type = "l", ylab = NA, xlab = "", ylim = c(-10,20), 
                            xaxt = "n", col = "dodgerblue", main = "Mean Temperature"))
    with(clim.Argentina, lines(y = Tav, x = c(1:12), col = "darkorange"))    
    with(clim.BritishColumbia, lines(y = Tav, x = c(1:12), col = "darkgreen"))    
    with(clim.NorthCarolina, lines(y = Tav, x = c(1:12), col = "darkorange"))    
        axis(side = 1, at = c(1:12), labels = month.abb, las = 2)

     
    
    
    
  
```

## Load example configuration data

```{r, results='hide', fig.align = 'center', fig.width = 5, fig.height = 5}



```


## Compare species

```{r species, results='hide', fig.align = 'center', fig.width = 5, fig.height = 5}
data("cfg.Pseudotsuga.menziesii")
data("cfg.Pinus.ponderosa")
pipo.run <- run_3pg(config = cfg.Pinus.ponderosa, 
                    climate = example.clim, 
                    output = FALSE, 
                    python_indexing = FALSE)
doug.run <- run_3pg(config = cfg.Pseudotsuga.menziesii, 
                    climate = example.clim, 
                    output = FALSE, 
                    python_indexing = FALSE)

example.clim$date <- apply(X = cbind(example.clim$Year, example.clim$Month), MARGIN = 1, FUN = function(x){paste(x[1], x[2], 15, sep = "-")})
example.clim$date <- strptime(x = example.clim$date, format = "%Y-%m-%d")

plot(tail(pipo.run$D13CTissue+1.5,100) ~ as.Date(tail(example.clim$date, 100)), type = "l", col = "dodgerblue", ylab = "d13C wood", xlab = "") # add 1.5 to get gfrom wood to cellulose
lines(tail(doug.run$D13CTissue+1.5,100) ~ as.Date(tail(example.clim$date, 100)), type = "l", col = "darkorange", ylab = "d13C wood", xlab = "")
```

## Compare species at sites

```{r speciessites, results='hide', fig.align = 'center', fig.width = 6, fig.height = 6}
data("clim.Argentina")
data("clim.BritishColumbia")
data("clim.NorthCarolina")
data("clim.WesternOR")

cfg.Pinus.ponderosa$TimeRange
  cfg.Pinus.ponderosa$TimeRange$EndAge <- 25
  cfg.Pinus.ponderosa$TimeRange$EndYear <- 1975

pipo.Arg.run <- run_3pg(config = cfg.Pinus.ponderosa, 
                    climate = clim.Argentina, 
                    output = FALSE, 
                    python_indexing = FALSE)
pipo.WOR.run <- run_3pg(config = cfg.Pinus.ponderosa, 
                    climate = clim.WesternOR, 
                    output = FALSE, 
                    python_indexing = FALSE)

pipo.Arg.run$D13CTissue <- pipo.Arg.run$D13CTissue + 1.5
pipo.WOR.run$D13CTissue <- pipo.WOR.run$D13CTissue + 1.5 

plot(tail(pipo.Arg.run$D13CTissue, 12) ~ 1, xaxt = "n", type = "l", col = "dodgerblue", ylab = "d13C wood", xlab = "")
  lines(tail(pipo.WOR.run$D13CTissue, 12) ~ 1, type = "l", col = "darkorange", ylab = "d13C wood", xlab = "")
  axis(side = 1, at = c(1:12), labels = month.abb)

data("cfg.Eucalyptus.globulus")
data("cfg.Picea.sitchensis")
data("cfg.Pinus.contorta")
data("cfg.Pinus.radiata")
data("cfg.Pinus.taeda")
```

## References

[Landsberg, J.J., and R.H. Waring. "A Generalised Model of Forest Productivity Using Simplified Concepts of Radiation-Use Efficiency, Carbon Balance and Partitioning." Forest Ecology and Management 95, no. 3 (August 1997): 209-28. https://doi.org/10.1016/S0378-1127(97)00026-1.](https://www.sciencedirect.com/science/article/pii/S0378112797000261)

[Wei, Liang, John D. Marshall, Jianwei Zhang, Hang Zhou, and Robert F. Powers. "3-PG Simulations of Young Ponderosa Pine Plantations under Varied Management Intensity: Why Do They Grow so Differently?" Forest Ecology and Management 313 (February 2014): 69-82. https://doi.org/10.1016/j.foreco.2013.10.035.](https://www.sciencedirect.com/science/article/pii/S0378112713007226)











